What's in a Name? Linear Temporal Logic Literally Represents Time Lines

### 2023-06-26 paper/main.tex
typos

\begin{abstract}
  Linear Temporal Logic (LTL) is arguably the most popular specification language for formal verification of safety-critical systems.
  However, LTL formulas can be unintuitive and error-prone for human practitioners to specify and validate. Meanwhile, drawing timelines remains one of the most popular methods for specifying and validating requriements for industrial system designs, such as in aerospace operational concepts. 
  Therefore, we provide a new timeline tool for visualizing LTL specifications as timelines, providing provably-correct, intuitive equivalents between these two specification formats.
  Our tool generates timeline visualizations by translating LTL formulas to intermediate representations as B\"{u}chi automata and then regular expressions, and finally simplifying and visualizing the expressions.
  We provide an algorithm for this visualization, a theoretical soundness analysis, and an implementation.
\end{abstract}

@@

\section{Theoretical Analyses} \label{sec:analysis}
In the following section, we prove the correctness of our translation pipeline. The currectness of the translation from LTL formulas to \Buchi automata is provided by SPOT~\cite{Dur22}. Below, we prove the correctness of (i) the translation from \Buchi automata to $\omega$-regular expressions, (ii) the ($\omega$-)regular expression rewrite rules we apply, and lastly, we outline the correctness of our timeline visualizations. 
\subsection{Correctness of Regex Translation}
\begin{lemma}
    % reduce_nfa does its input to output behaviour
    For any NFA $G$ and any state $v$ in $G$ that is neither a start state or a final state, \textbf{reduce\_nfa}$(G, v)$ preserves the regular language accepted by $G$
\end{lemma}
\begin{proof}
    Let $G'$ be the graph of $G$ post reduction by the application of \textbf{reduce\_nfa}$(G, v)$. We show that the trace of every path accepted by $G$ is also accepted by $G'$. Suppose a path accepted by $G$ does not pass through $v$, clearly this is true. Else, suppose it passes through $v$, since $v$ is an interior node, $v$ cannot be the first or last in the path. Thus, for every pass through $v$, let $u\neq v$ be the last node passed before entering $v$, and likewise $w$ be the first node after exiting $v$. We show that the regular language of sub-traces from $u$ to $w$ in $G$ (shown in \cref{fig:uvw-dfa}) is identical to that in $G'$.
    \begin{figure}[h!]
        \centering
        \begin{tikzpicture}
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (u) {$u$};
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (v) [right=of u] {$v$};
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (w) [right=of v] {$w$};
	\draw[thick] (u) edge["$r_{uv}$",-latex] (v) (v) edge["$r_{vv}$",-latex,loop above] (v) (v) edge["$r_{vw}$",-latex] (w);
        \end{tikzpicture}
 %       \includegraphics[scale=0.75]{img/uvw_dfa.pdf}
        \caption{A path from $u$ to $w$ in $G$}
        \label{fig:uvw-dfa}
    \end{figure}
    Suppose there is no loop at $v$, the path from $u$ to $w$ must simply be $u\xrightarrow{r_{uv}}v\xrightarrow{r_{vw}}w$, with the trace $r=r_{uv}r_{vw}$. If there is a loop edge $r_{vv}$, this edge can be traveled any number of times before exiting $v$, thus the regular expression is $r=r_{uv}r_{vv}^*r_{vw}$. This edge $u\xrightarrow{r}w$ is added in $G'$. Thus, for every path accepted by $G$, for every subpath entering and exiting $v$ from some $u$ to $w$, there exists a corresponding subpath from $u$ to $w$ in $G'$, and thus a corresponding path accepted by $G'$.
\end{proof}

@@

\subsection{Correctness of Timeline Visualization}\label{sec:correctness-timeline}
Inductively, we show how to construct each form of regular and $\omega$-regular expression into a timeline representation.
\paragraph{Regular expression}
\begin{enumerate}
  \item $A = \emptyset$. Visualized as no timeline. An empty regular expression signifies nothing is true at any time.
  \item $A = \epsilon$. Omitted in the visualization.
  \item $A = a$. Visualized as a single timeline node with $a$. A regular expression of an propositional logic formula signifies that the atomic proposition is true at the current time step.
  \item $A = A_1A_2$. Visualized as the timeline of $A_1$ followed by the timeline of $A_2$. A concatenation of two regular expressions signifies that the first regular expression must be true before the second regular expression can be true in the time step.
  \item $A = A_1 + A_2$. Visualized as the timelines of $A_1$ and $A_2$ in parallel. A union of two regular expressions signifies that either of the regular expressions can be true, which is represented by the parallel timelines.
  \item $A = A_1^*$. If $A_1$ is a propositional logic formula, then it is visualized as an egg shape node with caption ``repeats 0 - $\infty$''. If $A_1$ is a regular expression involving concatenation, star, etc, then it is visualized as the timeline of $A_1$ followed by a node with label ``$\cdots$'', and that timeline pattern again. A Kleene star of a regular expression signifies that the regular expression can be repeated any number of times, including 0 times, which matches the meaning of our representation of egg shape node and ``$\cdots$''  node as described in \cref{sec:timeline}.
\end{enumerate}
\paragraph{$\omega$-regular repression}
\begin{enumerate}
  \item $B = A^\omega$. Visualized as a grey box enclosed with the timeline of $A$. An $\omega$ of a regular expression signifies that the regular expression is to be repeated infinitely many times, which matches the meaning of our representation of grey box, because every time it reach the end of the grey box, it must reenter the same grey box again, hence infinite repeat.
  \item $B = AB$. Visualized as the timeline of $A$ followed by the timeline of $B$. A concatenation of a regular expression and an $\omega$-regular expression signifies that the first must be true before the second $\omega$-regular expression can be true in the following timesteps.
  \item $B = B + B$. Visualized as the timelines of $B$ and $B$ in parallel. A union of two $\omega$-regular expressions signifies that either of the $\omega$-regular expressions can be true, which is represented by the parallel timelines.
\end{enumerate}

>>>>

\begin{abstract}
  Linear Temporal Logic (LTL) is arguably the most popular specification language for formal verification of safety-critical systems.
  However, LTL formulas can be unintuitive and error-prone for human practitioners to specify and validate. Meanwhile, drawing timelines remains one of the most popular methods for specifying and validating requirements for industrial system designs, such as in aerospace operational concepts. 
  Therefore, we provide a new timeline tool for visualizing LTL specifications as timelines, providing provably-correct, intuitive equivalents between these two specification formats.
  Our tool generates timeline visualizations by translating LTL formulas to intermediate representations as B\"{u}chi automata and then regular expressions, and finally simplifying and visualizing the expressions.
  We provide an algorithm for this visualization, a theoretical soundness analysis, and an implementation.
\end{abstract}

@@

\section{Theoretical Analyses} \label{sec:analysis}
In the following section, we prove the correctness of our translation pipeline. The correctness of the translation from LTL formulas to \Buchi automata is provided by SPOT~\cite{Dur22}. Below, we prove the correctness of (i) the translation from \Buchi automata to $\omega$-regular expressions, (ii) the ($\omega$-)regular expression rewrite rules we apply, and lastly, we outline the correctness of our timeline visualizations. 
\subsection{Correctness of Regex Translation}
\begin{lemma}
    % reduce_nfa does its input to output behaviour
    For any NFA $G$ and any state $v$ in $G$ that is neither a start state or a final state, \textbf{reduce\_nfa}$(G, v)$ preserves the regular language accepted by $G$
\end{lemma}
\begin{proof}
    Let $G'$ be the graph of $G$ post reduction by the application of \textbf{reduce\_nfa}$(G, v)$. We show that the trace of every path accepted by $G$ is also accepted by $G'$. Suppose a path accepted by $G$ does not pass through $v$, clearly this is true. Else, suppose it passes through $v$, since $v$ is an interior node, $v$ cannot be the first or last in the path. Thus, for every pass through $v$, let $u\neq v$ be the last node passed before entering $v$, and likewise $w$ be the first node after exiting $v$. We show that the regular language of sub-traces from $u$ to $w$ in $G$ (shown in \cref{fig:uvw-dfa}) is identical to that in $G'$.
    \begin{figure}[h!]
        \centering
        \begin{tikzpicture}
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (u) {$u$};
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (v) [right=of u] {$v$};
	\node[draw, thick, shape=ellipse, style={minimum width=1.5cm, minimum height=1cm}] (w) [right=of v] {$w$};
	\draw[thick] (u) edge["$r_{uv}$",-latex] (v) (v) edge["$r_{vv}$",-latex,loop above] (v) (v) edge["$r_{vw}$",-latex] (w);
        \end{tikzpicture}
 %       \includegraphics[scale=0.75]{img/uvw_dfa.pdf}
        \caption{A path from $u$ to $w$ in $G$}
        \label{fig:uvw-dfa}
    \end{figure}
    Suppose there is no loop at $v$, the path from $u$ to $w$ must simply be $u\xrightarrow{r_{uv}}v\xrightarrow{r_{vw}}w$, with the trace $r=r_{uv}r_{vw}$. If there is a loop edge $r_{vv}$, this edge can be traveled any number of times before exiting $v$, thus the regular expression is $r=r_{uv}r_{vv}^*r_{vw}$. This edge $u\xrightarrow{r}w$ is added in $G'$. Thus, for every path accepted by $G$, for every subpath entering and exiting $v$ from some $u$ to $w$, there exists a corresponding subpath from $u$ to $w$ in $G'$, and thus a corresponding path accepted by $G'$.
\end{proof}

@@

\subsection{Correctness of Timeline Visualization}\label{sec:correctness-timeline}
Inductively, we show how to construct each form of regular and $\omega$-regular expression into a timeline representation.
\paragraph{Regular expression}
\begin{enumerate}
  \item $A = \emptyset$. Visualized as no timeline. An empty regular expression signifies nothing is true at any time.
  \item $A = \epsilon$. Omitted in the visualization.
  \item $A = a$. Visualized as a single timeline node with $a$. A regular expression of an propositional logic formula signifies that the atomic proposition is true at the current time step.
  \item $A = A_1A_2$. Visualized as the timeline of $A_1$ followed by the timeline of $A_2$. A concatenation of two regular expressions signifies that the first regular expression must be true before the second regular expression can be true in the time step.
  \item $A = A_1 + A_2$. Visualized as the timelines of $A_1$ and $A_2$ in parallel. A union of two regular expressions signifies that either of the regular expressions can be true, which is represented by the parallel timelines.
  \item $A = A_1^*$. If $A_1$ is a propositional logic formula, then it is visualized as an egg shape node with caption ``repeats 0 - $\infty$''. If $A_1$ is a regular expression involving concatenation, star, etc, then it is visualized as the timeline of $A_1$ followed by a node with label ``$\cdots$'', and that timeline pattern again. A Kleene star of a regular expression signifies that the regular expression can be repeated any number of times, including 0 times, which matches the meaning of our representation of egg shape node and ``$\cdots$''  node as described in \cref{sec:timeline}.
\end{enumerate}
\paragraph{$\omega$-regular repression}
\begin{enumerate}
  \item $B = A^\omega$. Visualized as a grey box enclosed with the timeline of $A$. An $\omega$ of a regular expression signifies that the regular expression is to be repeated infinitely many times, which matches the meaning of our representation of grey box, because every time it reach the end of the grey box, it must reenter the same grey box again, hence infinite repeat.
  \item $B = AB$. Visualized as the timeline of $A$ followed by the timeline of $B$. A concatenation of a regular expression and an $\omega$-regular expression signifies that the first must be true before the second $\omega$-regular expression can be true in the following time steps.
  \item $B = B + B$. Visualized as the timelines of $B$ and $B$ in parallel. A union of two $\omega$-regular expressions signifies that either of the $\omega$-regular expressions can be true, which is represented by the parallel timelines.
\end{enumerate}
